---
title: "Stroke Risk Prediction: Decision Tree Model Comparison"
author: "Tori Green"
date: "2025-06-17"
output: html_document
---

## 📖 Overview
Stroke is a leading cause of death and disability worldwide. This R Markdown provides all code and explanations for predicting stroke risk using decision tree models. It’s structured for GitHub with clear comments, narrative, and reproducible code.

---

## 🎯 Objective
- **Goal:** Develop and compare decision tree models for predicting stroke risk.
- **Key Questions:**
    - Which health and lifestyle factors most influence stroke?
    - How do different tree setups affect accuracy and sensitivity?

---

## 📂 Setup
```{r}
# Load required libraries
library(readxl)
library(dplyr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(pROC)
library(corrplot)
library(ggplot2)

# Read dataset
stroke <- read_excel("stroke_dataset.xlsx")

# View initial structure
str(stroke)
summary(stroke)
```
---

## 🔧 Data Preparation
```{r}
# Remove unnecessary ID column
stroke$id <- NULL

# Convert variables to factors where appropriate
stroke <- stroke %>% mutate(
  stroke = as.factor(stroke),
  gender = as.factor(gender),
  ever_married = as.factor(ever_married),
  work_type = as.factor(work_type),
  Residence_type = as.factor(Residence_type),
  smoking_status = as.factor(smoking_status),
  hypertension = as.factor(hypertension),
  heart_disease = as.factor(heart_disease),
  bmi = as.numeric(bmi)
)

# Check for missing values
colSums(is.na(stroke))

# Impute missing BMI values with mean
stroke$bmi[is.na(stroke$bmi)] <- mean(stroke$bmi, na.rm = TRUE)

# Check class balance
table(stroke$stroke)
round(prop.table(table(stroke$stroke)) * 100, 2)
```
---

## 📊 Correlation Analysis
```{r}
# Convert factors to numeric for correlation heatmap
stroke_numeric <- stroke %>% mutate(across(where(is.factor), as.numeric))
cor_matrix <- cor(stroke_numeric, use = "complete.obs")

# Correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", addCoef.col = "black", number.cex = 0.7)
```
**Comment:** Age, glucose level, and hypertension show strongest relationships with stroke.
---

## 📂 Train/Test Split
```{r}
set.seed(123)
train_index <- sample(nrow(stroke), floor(0.8 * nrow(stroke)))
train <- stroke[train_index, ]
test <- stroke[-train_index, ]

# Set training control with upsampling for class imbalance
ctrl <- trainControl(method = "cv", number = 10, sampling = "up")
```
---

## 🌳 Model 1: Full Decision Tree
```{r}
set.seed(123)
tree_model <- train(
  stroke ~ ., data = train,
  method = "rpart",
  trControl = ctrl,
  tuneGrid = expand.grid(cp = seq(0.005, 0.05, by = 0.005)),
  control = rpart.control(maxdepth = 4, minsplit = 20, cp = 0.005)
)

# Variable importance
var_imp <- varImp(tree_model, scale = FALSE)
plot(var_imp)

# Extract top variables
importance <- var_imp$importance
par(mar = c(14, 4, 4, 2))
barplot(sort(importance[, 1], decreasing = TRUE),
        main = "Variable Importance",
        col = "lightblue",
        las = 2,
        names.arg = rownames(importance))

top_vars <- head(importance[order(-importance[,1]), , drop = FALSE], 5)
top_vars
```
---

## 📊 Model 1 Evaluation
```{r}
pred_probs <- predict(tree_model, newdata = test, type = "prob")
pred_class <- predict(tree_model, newdata = test)
conf_mat <- confusionMatrix(pred_class, test$stroke)
conf_mat

# ROC and AUC
roc_obj <- roc(response = test$stroke, predictor = pred_probs[, "1"])
roc_obj
```
---

## 🌳 Model 2: Simplified Tree
```{r}
set.seed(123)
tree_model2 <- train(
  stroke ~ age + avg_glucose_level + ever_married + hypertension + heart_disease + bmi,
  data = train,
  method = "rpart",
  trControl = ctrl,
  tuneGrid = expand.grid(cp = seq(0.005, 0.05, by = 0.005)),
  control = rpart.control(maxdepth = 4, minsplit = 20, cp = 0.005)
)

# Model 2 evaluation
pred_probs2 <- predict(tree_model2, newdata = test, type = "prob")
pred_class2 <- predict(tree_model2, newdata = test)
conf_mat2 <- confusionMatrix(pred_class2, test$stroke)
conf_mat2
roc_obj2 <- roc(response = test$stroke, predictor = pred_probs2[, "1"])
roc_obj2
```
---

## 📈 Model Comparison
```{r}
model_summary <- data.frame(
    Model = c("Decision Tree Model", "Decision Tree Model2"),
    Kappa = c(0.1918, 0.1726),
    Accuracy = c(0.7466, 0.7456),
    Sensitivity = c(0.7412, 0.7443),
    Specificity = c(0.8393, 0.7679),
    AUC = c(0.8181, 0.8347)
)
model_summary
```
**Comment:** Model 2 shows slightly higher AUC but lower specificity.
---

## 📉 ROC Curve & Final Tree Plot
```{r}
plot(roc_obj2, col = "#2C77B2", lwd = 3, main = "ROC Curve - Stroke Prediction")
legend("bottomright", legend = paste("AUC =", round(auc(roc_obj2), 3)), col = "#2C77B2", lwd = 2)

# Decision tree visualization
rpart.plot(tree_model2$finalModel,
           type = 2,
           extra = 104,
           fallen.leaves = TRUE,
           cex = 0.8,
           box.palette = "GnBu")
```
---

## 🩺 Rule Extraction
```{r}
# Extract rules
rules <- asRules(tree_model2$finalModel)
rules

# Print top risk indicators
cat("\nTop Risk Indicators:\n",
    paste("- ", rownames(top_vars), ": Monitor closely.", sep = "", collapse = "\n"))
```
---

## 💡 Insights
- Age, glucose level, and hypertension were top predictors.
- Model 2 was simpler and slightly improved AUC.
- Upsampling improved stroke detection.

---

## 🚀 Next Steps
- Consider ensemble models for better accuracy.
- Implement SMOTE or cost-sensitive learning.
- Wrap model into a Shiny app for clinicians.

---

## 🛠 Tools Used
- caret – Model training & CV
- rpart & rpart.plot – Decision trees
- pROC – ROC/AUC
- corrplot – Correlation heatmap
- rattle – Rule extraction

---


